Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using Multiple Automatically Generated Templates

### 2013-09-02 paper.Rnw
Abstract complete, todo's removed, and abbrv commands applied

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has 
%  - add MAGeT brain diagram, and remove pseudocode
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\end{abstract}

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 tests hippocampal subfield segmentation quality. The last two
experiments assess the validity of the \mb algorithm when applied to different
diseases: first episode schizophrenia (Experiment 3), and Alzheimer's disease
(Experiment 4). Additionally, in Experiment 4, we compare \mb segmentations with
those of well-known automated and manual methods and assess segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and SNT \gs segmentations from the ADNI1 dataset \citep{Hsu2002}. This
form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

@@

\paragraph{Winterburn Atlases dataset} 
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $300\mu$ isotropic T1-weighted MR images. The
segmentations include subfield segmentations for the cornu ammonis (CA) 1; CA2
and CA3; CA4 and dentate gyrus; subiculum; and strata radiatum (SR, strata
lacunosum (SL), and strata moleculare (SM). Subjects in the Winterburn atlases
range in age from 29-57 years (mean age of 37), and include two males and three
females.

@@

\paragraph{ADNI1 dataset revisited} 
The {\em \adnidataset} standardized dataset contains $1919$ images in total.
SNT and FreeSurfer hippocampal volumes for a subset of images were provided by
ADNI, along with quality control data for each \freesurfer segmentation
(guidelines described in \citep{Hartig2010}).  Clinical and demographic data for
the entire \adnidataset dataset are shown in Table
\ref{tab:ADNI1-Dataset-Demographics}.

\paragraph{Experiment details} 
\mb was configured with an atlas library composed of the five Winterburn T1
atlases (described in Experiment 2), and size of template library,
registration method, and label fusion method were determined by the optimal
settings found in Experiment 1.  The FSL FIRST segmentation method was used via
the \verb+run_first_all+ script according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}.  All images
in the \adnidataset dataset were segmented by both methods.

One author (MP) performed visual quality inspection for MAGeT and FSL
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by $10mm$ or greater in three or more slices then
the segmentation did not pass).  Only images meeting the conditions of having
segmentations from all methods (SNT, MAPER, FreeSurfer, FSL, and \mb) and also
passing quality control inspection were included in the analysis.

@@

In general, across hippocampal subregions the percent error in volume of \mb
segmentations compares favourably to the error resulting from image resampling
(Figure \ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4/Dentate, and
SR/SL/SM subregions all show near or smaller percent errors. The Subiculum and
CA2/CA3 subregions show distinctly larger error than is found through
resampling.

@@

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly chosen
subject images (7 from each disease class) and majority vote label fusion.
The entire \adnidataset dataset was segmented by \mb, and the resulting volumes
compared with those obtained by expert semi-automated segmentation (SNT), and
three other automated segmentation techniques: MAPER, FreeSurfer, and FSL.
Table \ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total
count of segmentations available, including a count of those which have failed
a quality control inspection.  A total of \Sexpr{length(means.complete$RID)}
images are included in the following analysis, having met quality criteria and
having segmentations from every method.

@@

To investigate the level of agreement with SNT \gs hippocampal
volumes, we constructed Bland-Altman plots for each method (Figure
\ref{fig:ADNI-Bland-Altman}).  As \citet{Bland1986} noted, high correlation
amongst measures of the same quantity does not necessarily imply agreement (as
correlation can be driven by a large range in true values, for instance).  All
methods show an obvious proportional bias: FreeSurfer and FSL markedly
under-estimate smaller hippocampi and over-estimate large hippocampi, whereas
MAPER and \mb show a reverse, conservative bias (Figure
\ref{fig:ADNI-Bland-Altman}).  Additionally, all methods show a fixed volume
bias, with FreeSurfer and FSL most dramatically over-estimating hippocampal
volume by $2600 mm^3$ and $2800 mm^3$ on average, respectively, and MAPER and
\mb within $250 mm^3$ on average.

@@

qplot(DX,value,data=m2,
      colour=Method,geom="boxplot") + 
      #geom_text(aes(y=-Inf,x=c(.7,.85,1,1.15, 1.7,1.85,2,2.15, 2.7,2.85,3,3.15), vjust=-5, label=round(pearson,2)), 
      #          colour = 'black', size=3, data=correlations) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Hippocampal volume (", mm^3, ")"))) + 
      opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset
dataset by FreeSurfer, FSL, MAPER, \mb, and SNT \gs vs. disease
category.}
  \label{fig:ADNI-volumes-boxplot}
\end{figure}

@@

ggplot(data=melted, aes(x=SNT, y=volume, colour=Method)) + 
    geom_point(size=1) + 
    geom_smooth(method="lm") + 
    xlab(expression("SNT \\gs mean hippocampus volume " (mm^3))) + 
    ylab(expression("Automated mean hippocampal volume " (mm^3))) +
    annotate("text", y = c(2000,1500,1000,500), x=Inf-100, size=2.5,
              label = c(paste("FS r =", round(cor(df$SNT, df$FS),2)),
                        paste("FSL r =", round(cor(df$SNT, df$FSL),2)),
                        paste("MAPER r =", round(cor(df$SNT, df$MAPER),2)),
                        paste("MAGeT r =", round(cor(df$SNT, df$MAGeT),2))),
              hjust=1.1,vjust=1.5) +
    opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset
dataset by each of the four automated methods investigated (FreeSurfer (FS),
FSL, MAPER, \mb) vs. SNT \gs. Linear fit lines and Pearson
correlations are shown for each method.}
  \label{fig:ADNI-volumes-plot}
\end{figure}

@@

names(melted)[4] = "Diagnosis"
ggplot(melted, aes(x= mean, y = diff, colour=Diagnosis)) + 
  facet_wrap( ~ method, nrow=2, scales="fixed") + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits,
            hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of SNT \\gs and automated hippocampal volume ' (mm^3))) + 
  ylab(expression('Difference in manual and automated hippocampal volume ' (mm^3))) + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@
  \caption{Bland-Altman plots comparing subject mean hippocampal volume as
measured in the \adnidataset dataset by SNT \gs segmentation and each of the
four automated methods investigated (FreeSurfer, FSL, MAPER, \mb). The overall
mean difference in volume, and limits of agreement ($\pm 1.96SD$) are shown by
dashed horizontal lines. Linear fit lines are shown for each diagnosis group.
Note, points below the mean difference indicate overestimation of the volume
with respect to the \gs rating, and vice versa. }

@@

In this manuscript we have presented the implementation and validation of the
\mb framework -- a methodology that requires very few input atlases in
order to provide accurate and reliable segmentations.  Experiment 1 establishes
\mb as effective, and as comparable to multi-atlas segmentation. This experiment
also rigorously characterizes the behaviour of \mb whole hippocampal
segmentation under various parameter settings allowing us to choose an optimal
setting for subsequent experiments. Experiment 2 demonstrates the accuracy of
\mb hippocampal subfield identification despite contrast and resolution
limitations in standard T1-weighted image volumes.  Experiments 2 and 3 validate
\mb whole hippocampal segmentation accuracy and consistency on populations with
different ageing and neuropsychiatric characteristics.  Furthermore, taken
together, these experiments demonstrate that algorithmic performance is not
dependent on a single definition of the hippocampus but is effective with
differing hippocampal definitions \citep{Winterburn2013,Pruessner2000,Hsu2002}.

@@

Another concern is the moderate-to-small improvement observed in \mb in
comparison to multi-atlas segmentation when using the same number of atlases.
The actual benefit in using \mb is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo leave-one-out cross-validation
that we present in Experiment 1 is amongst one of the most stringent performed
in the multi-atlas/segmentation literature. To the best of our knowledge, other
groups using ADNI data for validation do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}), aside from
\citep{Wang2011} who perform a similarily rigorous validation as us). The
thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice (provided the input atlases are consistently segmented).

@@

To the best of our knowledge, no other groups have validated their work using
multiple atlas segmentation protocols, different acquisitions, and disease
populations in order to demonstrate the robustness of their technique.  This is
one of the clear strengths of this work.  Furthermore, unlike some of the
algorithms mentioned, our implementation does not require retuning for new
populations or datasets as it inherently models the variability of the dataset
through the template library.  However it should be noted that the increased
accuracy that follows increasing the number of atlases and templates comes at an
increased computational cost ($O(log(n))$), as previously mentioned in
other work \citep{Heckemann2006}. 
\todo[inline]{time complexity seems suspicious}

Among the automated segmentation methods we compared in this paper
(\freesurfer, \maper, \fsl-FIRST), we find extremely variable
performance of all methods.  With the exception of \fsl all methods correlate
well with the SNT volumes provided in the ADNI database.  However, \freesurfer
and \fsl provide radically different definitions of the size of the hippocampus
in comparison to the other methods.  Further, when estimating bias of these
methods relative to SNT \gs hippocampal volumes we see that large hippocampi are
over estimated while small hippocampi are under estimated. By comparison, \mb
and MAPER are far more conservative in volume estimation, suggesting these
methods may be better suited for estimating true-positives, especially in
neurodegenerative disease subjects featuring smaller overall hippocampi.
However, in this analysis we have only compared methods by total hippocampal
volume, and so more work is needed to understand the full extent to which these
methods differ.

Finally, we have also demonstrated that our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data.  Subfield segmentation is a burgeoning topic in the
literature although very few automated methods are available for the
segmentation of 3T data \citep{Yushkevich2009,VanLeemput2009}. However, the
FreeSurfer implementation of subfield segmentation \todo{complete}. While
recent work demonstrates that subfield segmentations can be used for
classification of AD, MCI, and CN, there has been no explicit validation of the
methodology based on accuracy or precision.  Although the initial hippocampal
subfield segmentations from the Yushkevich group have been demonstrated to work
on the ADNI population, there has also been no validation of their work.  In
addition, their work requires some manual initialization to properly function.
Our work demonstrates that we can reliably identify the CA1, subiculum, and
CA4/dentate with only modest amounts of error.  The fact that CA2/CA3 and
molecular layers cannot be reasonably identified should not be surprising as
these are extremely thin and spatially convoluted regions that originally
required high-resolution MRI for identification.  It is likely that the extents of
these regions are well below the resolution offered by standard T1-weighted
images and, in fact, many manual segmentation methodologies do not attempt to
parse these regions either \citep{Wisse2012, Mueller2009}.

>>>>

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has 
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

\begin{abstract}

\textbf{Introduction:} Neuroimaging research often relies on automated anatomical
segmentations of MR images of the brain. Current multi-atlas based approaches
provide accurate segmentations of brain images by propagating manually derived
segmentations of specific neuroanatomical structures to unlabelled data. These
approaches often rely on a large number of such manually segmented atlases (e.g.
30-80) that take significant time and expertise to produce. We present an
algorithm, \mb, for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
multi-atlas approaches.\\
\textbf{Method:} \mb works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 69 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb when applied to
Schizophrenia and Alzheimer's disease populations, and \mb segmentations are
compared with existing automated methods (\fsl, \freesurfer, \maper) and biases are
explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo
cross-validation on 69 subjects relative to semi-automated \gs segmentations,
and shows significantly lower variability in DSC than multi-atlas segmentation.
In a LOOCV, \mb reproduces the volumes of the CA1, CA4/DG, and SR/SL/SM
hippocampal subregions with a percent error in volume that is at or lower than
that produced by image resampling. \mb produces hippocampal volumes in a
Schizophrenia patient population that are highly correlated with expert manual
segmentation volumes (Pearson $r = 0.877, t = 16.244, p < 0.001$). Compared to
\fsl and \freesurfer, \mb shows much smaller fixed volume bias (within $250
mm^3$ on average) to \gs segmentations, as well as a conservative, rather than
exaggerated, proportional volume bias.\\
\textbf{Conclusion:} We demonstrate that \mb produces accurate hippocampal
segmentations using only 5 atlases over different hippocampal definitions,
disease populations, and acquisition types, as well as showing that accurate
identification of the hippocampal subfields is possible.
\end{abstract}

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compare \mb segmentations with those of
well-known automated methods and assess segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and SNT \gs segmentations from the ADNI dataset \citep{Hsu2002}. This
form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

@@

\paragraph{Winterburn Atlases dataset} 
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $300\mu$ isotropic T1-weighted MR images. The
segmentations include subfield segmentations for the cornu ammonis (CA) 1; CA2
and CA3; CA4 and dentate gyrus; subiculum; and strata radiatum (SR), strata
lacunosum (SL), and strata moleculare (SM). Subjects in the Winterburn atlases
range in age from 29-57 years (mean age of 37), and include two males and three
females.

@@

\paragraph{ADNI1 dataset revisited} 
The {\em \adnidataset} standardized dataset contains $1919$ images in total.
SNT and \freesurfer hippocampal volumes for a subset of images were provided by
ADNI, along with quality control data for each \freesurfer segmentation
(guidelines described in \citep{Hartig2010}).  Clinical and demographic data for
the entire \adnidataset dataset are shown in Table
\ref{tab:ADNI1-Dataset-Demographics}.

\paragraph{Experiment details} 
\mb was configured with an atlas library composed of the five Winterburn T1
atlases (described in Experiment 2), and size of template library,
registration method, and label fusion method were determined by the optimal
settings found in Experiment 1.  The \freesurfer segmentation method was used via
the \verb+run_first_all+ script according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}.  All images
in the \adnidataset dataset were segmented by both methods.

One author (MP) performed visual quality inspection for \mb and \freesurfer 
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by $10mm$ or greater in three or more slices then
the segmentation did not pass).  Only images meeting the conditions of having
segmentations from all methods (SNT, \maper, \freesurfer, \fsl, and \mb) and also
passing quality control inspection were included in the analysis.

@@

In general, across hippocampal subregions the percent error in volume of \mb
segmentations (relative to the full-resolution Winterburn atlas segmentation)
compares favourably to the error resulting from image resampling (Figure
\ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4/DG, and SR/SL/SM
subregions all show a percent error in volume that is at or lower than
resampling error. The Subiculum and CA2/CA3 subregions show distinctly larger
percent error in volume than is found through resampling.

@@

Based on the results from Experiment 1, in this experiment \mb was configured
with a template library of 21 randomly chosen subject images (7 from each
disease class) and majority vote label fusion.  The entire \adnidataset dataset
was segmented by \mb, and the resulting volumes compared with those obtained by
expert semi-automated segmentation (SNT), and three other automated segmentation
techniques: \maper, \freesurfer, and \fsl. Table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total count of
segmentations available, including a count of those which have failed a quality
control inspection.  A total of \Sexpr{length(means.complete$RID)} images are
included in the following analysis, having met quality criteria and having
segmentations from every method.

@@

To investigate the level of agreement with SNT \gs hippocampal
volumes, we constructed Bland-Altman plots for each method (Figure
\ref{fig:ADNI-Bland-Altman}).  As \citet{Bland1986} noted, high correlation
amongst measures of the same quantity does not necessarily imply agreement (as
correlation can be driven by a large range in true values, for instance).  All
methods show an obvious proportional bias: \freesurfer and \fsl markedly
under-estimate smaller hippocampi and over-estimate large hippocampi, whereas
\maper and \mb show a reverse, conservative bias (Figure
\ref{fig:ADNI-Bland-Altman}).  Additionally, all methods show a fixed volume
bias, with \freesurfer and \fsl most dramatically over-estimating hippocampal
volume by $2600 mm^3$ and $2800 mm^3$ on average, respectively, and \maper and
\mb within $250 mm^3$ on average.

@@

qplot(DX,value,data=m2,
      colour=Method,geom="boxplot") + 
      #geom_text(aes(y=-Inf,x=c(.7,.85,1,1.15, 1.7,1.85,2,2.15, 2.7,2.85,3,3.15), vjust=-5, label=round(pearson,2)), 
      #          colour = 'black', size=3, data=correlations) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Hippocampal volume (", mm^3, ")"))) + 
      opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset
dataset by \freesurfer, \fsl, \maper, \mb, and SNT \gs vs. disease
category.}
  \label{fig:ADNI-volumes-boxplot}
\end{figure}

@@

ggplot(data=melted, aes(x=SNT, y=volume, colour=Method)) + 
    geom_point(size=1) + 
    geom_smooth(method="lm") + 
    xlab(expression("SNT \\gs mean hippocampus volume " (mm^3))) + 
    ylab(expression("Automated mean hippocampal volume " (mm^3))) +
    annotate("text", y = c(2000,1500,1000,500), x=Inf-100, size=2.5,
              label = c(paste("FS r =", round(cor(df$SNT, df$FS),2)),
                        paste("FSL r =", round(cor(df$SNT, df$FSL),2)),
                        paste("MAPER r =", round(cor(df$SNT, df$MAPER),2)),
                        paste("MAGeT r =", round(cor(df$SNT, df$MAGeT),2))),
              hjust=1.1,vjust=1.5) +
    opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset
dataset by each of the four automated methods investigated (\freesurfer (FS), \fsl,
\maper, \mb) vs. SNT \gs. Linear fit lines and Pearson correlations are shown
for each method.}
  \label{fig:ADNI-volumes-plot}
\end{figure}

@@

names(melted)[4] = "Diagnosis"
ggplot(melted, aes(x= mean, y = diff, colour=Diagnosis)) + 
  facet_wrap( ~ method, nrow=2, scales="fixed") + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits,
            hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of SNT \\gs and automated hippocampal volume ' (mm^3))) + 
  ylab(expression('Difference in manual and automated hippocampal volume ' (mm^3))) + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@
  \caption{Bland-Altman plots comparing subject mean hippocampal volume as
measured in the \adnidataset dataset by SNT \gs segmentation and each of the
four automated methods investigated (\freesurfer, \fsl, \maper, \mb). The overall
mean difference in volume, and limits of agreement ($\pm 1.96SD$) are shown by
dashed horizontal lines. Linear fit lines are shown for each diagnosis group.
Note, points below the mean difference indicate overestimation of the volume
with respect to the \gs rating, and vice versa. }

@@

In this manuscript we have presented the implementation and validation of the
\mb framework -- a methodology that requires very few input atlases in
order to provide accurate and reliable segmentations.  Experiment 1 establishes
\mb as effective and as comparable to multi-atlas segmentation. This experiment
also rigorously characterizes the behaviour of \mb whole hippocampal
segmentation under various parameter settings, allowing us to choose an optimal
setting for subsequent experiments. Experiment 2 demonstrates the accuracy of
\mb hippocampal subfield identification despite contrast and resolution
limitations in standard T1-weighted image volumes.  Experiments 2 and 3 validate
\mb whole hippocampal segmentation accuracy and consistency on populations with
different ageing and neuropsychiatric characteristics.  Furthermore, taken
together, these experiments demonstrate that algorithmic performance is not
dependent on a single definition of the hippocampus but is effective with
differing hippocampal definitions \citep{Winterburn2013,Pruessner2000,Hsu2002}.

@@

Another concern is the moderate-to-small improvement observed in \mb in
comparison to multi-atlas segmentation when using the same number of atlases.
The actual benefit in using \mb is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo leave-one-out cross-validation
that we present in Experiment 1 is amongst one of the most stringent performed
in the multi-atlas/segmentation literature. To the best of our knowledge, other
groups using ADNI data for validation do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}), aside from
\citep{Wang2011} who perform a similarily rigorous validation as us. The
thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice (provided the input atlases are consistently segmented).

@@

To the best of our knowledge, no other groups have validated their work using
multiple atlas segmentation protocols, different acquisitions, and disease
populations in order to demonstrate the robustness of their technique.  This is
one of the clear strengths of this work.  Furthermore, unlike some of the
algorithms mentioned, our implementation does not require retuning for new
populations or datasets as it inherently models the variability of the dataset
through the template library.  However it should be noted that the increased
accuracy that follows increasing the number of atlases and templates comes at an
increased computational cost ($O(log(n))$), as previously mentioned in
other work \citep{Heckemann2006}.

Among the automated segmentation methods we compared in this paper
(\freesurfer, \maper, \fsl-FIRST), we find extremely variable
performance of all methods.  With the exception of \fsl all methods correlate
well with the SNT volumes provided in the ADNI database.  However, \freesurfer
and \fsl provide radically different definitions of the size of the hippocampus
in comparison to the other methods.  Further, when estimating bias of these
methods relative to SNT \gs hippocampal volumes we see that large hippocampi are
over estimated while small hippocampi are under estimated. By comparison, \mb
and \maper are far more conservative in volume estimation, suggesting these
methods may be better suited for estimating true-positives, especially in
neurodegenerative disease subjects featuring smaller overall hippocampi.
However, in this analysis we have only compared methods by total hippocampal
volume, and so more work is needed to understand the full extent to which these
methods differ.

Finally, we have also demonstrated that our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data.  Subfield segmentation is a burgeoning topic in the
literature although very few automated methods are available for the
segmentation of 3T data \citep{Yushkevich2009,VanLeemput2009}. While
recent work demonstrates that subfield segmentations can be used for
classification of AD, MCI, and CN, there has been no explicit validation of the
methodology based on accuracy or precision.  Although the initial hippocampal
subfield segmentations from the Yushkevich group have been demonstrated to work
on the ADNI population, there has also been no validation of their work. In
addition, their work requires some manual initialization to properly function.
Our work demonstrates that we can reliably identify the CA1, subiculum, and
CA4/dentate with only modest amounts of error.  The fact that CA2/CA3 and
molecular layers cannot be reasonably identified should not be surprising as
these are extremely thin and spatially convoluted regions that originally
required high-resolution MRI for identification.  It is likely that the extents of
these regions are well below the resolution offered by standard T1-weighted
images and, in fact, many manual segmentation methodologies do not attempt to
parse these regions either \citep{Wisse2012, Mueller2009}.
